# Exploratory Data Analysis on US Police Shootings

```{r}
library(ggplot2)
library(dplyr)
```

**The Washington Post,an American daily newspaper began in 2015 to log person shot and killed by an on-duty police officer in the United States.**

> After **Michael Brown**, an unarmed Black man, was killed in 2014 by police, a Washington Post investigation found that data reported to the FBI on fatal police shootings was undercounted by more than **half**. That gap has widened in recent years. By 2021, only a **third** of departments' fatal shootings appeared in the **FBI database**.

**This data set contains the list of victims from 2015 to 2020.**
#E:\Semester 1\US_Police_Shootings
# DATA DESCRIPTION

```{r}
#setwd("C:/Semester 1/US_Police_Shootings/")
dataset = read.csv("USPolice.csv")
View(dataset)
```

```{r}

#View(dataset)
```

## SUMMARY

### **Columns and their type in the data set**

```{r}
#dataset <- subset(dataset, select = -c(name)

dataset <- transform(dataset, date  = as.Date(date))

my_colors = c("#a53860", "#FFD166", "#60712f", "#118AB2", "#073B4C", "#d8a47f")

print(summary(dataset))
```

# Demographic Analysis

## RACE ABBREVIATION

```{r}
race_data <- data.frame(
  race = c("Asian","Black","Hispanic","Native American","Others","White" ),
  Abbreviation = c("A","B","H","N","O","W")
)

print(race_data)

```

```{r}
# Calculate the percentage of each gender
percentage_data <- dataset %>%
  group_by(gender) %>%
  summarise(count = n(), percentage = (n() / nrow(dataset)) * 100)

# Create the pie chart
p1 <- ggplot(percentage_data, aes(x = "", y = percentage, fill = gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  labs(
    title = "Victims Based on Gender",
    fill = "Gender"
  ) +
  scale_fill_manual(values = my_colors) +
  geom_text(aes(label = paste0(round(percentage), "%")), position = position_stack(vjust = 0.5)) +
  theme_void()



p2 <- dataset %>%
  ggplot(aes(race))+
  geom_bar( fill=my_colors)+
  theme(axis.text.x=element_text(angle=0, hjust=1))+
  theme(text=element_text(size=10))
  ggtitle("Race Distribution")
  

p1
p2

```

```{r}
# Calculate the counts for each armed weapon
armed_counts <- table(dataset$armed)

# Sort and select the top 5 armed weapons
top_weapons <- head(sort(armed_counts, decreasing = TRUE), 5)

# Combine the rest as "other"
other_count <- sum(armed_counts[!(names(armed_counts) %in% names(top_weapons))])

# Create a data frame for the pie chart
pie_data <- data.frame(
  weapon = c(names(top_weapons), "Other"),
  count = c(top_weapons, other_count)
)

# Calculate percentages
pie_data$percentage <- pie_data$count / sum(pie_data$count) * 100

# Create the pie chart


p0<-ggplot(pie_data, aes(x = "", y = percentage, fill = weapon)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(round(percentage), "%")), position = position_stack(vjust = 0.5)) +
  theme_void() +
  scale_fill_manual(values = my_colors) +
  ggtitle("Top 5 Victims' Weapons and Unarmed")



# Calculate the percentage variable first
percentage_data <- dataset %>%
  group_by(flee) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Create the ggplot
p1 <- ggplot(percentage_data, aes(x = "", y = percentage, fill = flee)) +
  geom_col(color = "black") +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  ggtitle("Victims Fleeing") +
  scale_fill_manual(values = my_colors) +
  labs(title = "Victims Fleeing", fill = "With")

    
  
# Calculate the percentage variable first
percentage_data <- dataset %>%
  group_by(manner_of_death) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Create the ggplot
p2 <- ggplot(percentage_data, aes(x = "", y = percentage, fill = manner_of_death)) +
  geom_col(color = "black") +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  ggtitle("Manner of Death") +
  scale_fill_manual(values = my_colors) +
  labs(title = "Manner of Death", fill = "Manner")





# Calculate the percentage variable first
percentage_data <- dataset %>%
  group_by(signs_of_mental_illness) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Create the ggplot
p3 <- ggplot(percentage_data, aes(x = "", y = percentage, fill = signs_of_mental_illness)) +
  geom_col(color = "black") +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  ggtitle("Mental Illness") +
  scale_fill_manual(values = my_colors) +
  labs(title = "Mental Illness", fill = "Has Mental Illness")




# Calculate the percentage variable first
percentage_data <- dataset %>%
  group_by(threat_level) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Create the ggplot
p4 <- ggplot(percentage_data, aes(x = "", y = percentage, fill = threat_level)) +
  geom_col(color = "black") +
  coord_polar(theta = "y", start = 0) +
  geom_text(aes(label = paste0(round(percentage), "%")),
            position = position_stack(vjust = 0.5)) +
  theme_void() +
  ggtitle("Threat Level") +
  scale_fill_manual(values = my_colors) +
  labs(title = "Threat Level", fill = "Threat level Posed:")


p0
p1
p2
p3
p4
```

## **TWO-LETTER STATE AND TERRITORY ABBREVIATIONS**

```{r}
# Create a data frame with state names and abbreviations
state_data <- data.frame(
  State = c(
    "Alabama", "Alaska", "Arizona", "Arkansas", "American Samoa",
    "California", "Colorado", "Connecticut", "Delaware", "District of Columbia",
    "Florida", "Georgia", "Guam", "Hawaii", "Idaho", "Illinois", "Indiana",
    "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts",
    "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana", "Nebraska",
    "Nevada", "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina",
    "North Dakota", "Northern Mariana Islands", "Ohio", "Oklahoma", "Oregon",
    "Pennsylvania", "Puerto Rico", "Rhode Island", "South Carolina", "South Dakota",
    "Tennessee", "Texas", "Trust Territories", "Utah", "Vermont", "Virgin Islands",
    "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"
  ),
  Abbreviation = c(
    "AL", "AK", "AZ", "AR", "AS",
    "CA", "CO", "CT", "DE", "DC",
    "FL", "GA", "GU", "HI", "ID", "IL", "IN",
    "IA", "KS", "KY", "LA", "ME", "MD", "MA",
    "MI", "MN", "MS", "MO", "MT", "NE", "NV",
    "NH", "NJ", "NM", "NY", "NC", "ND", "MP",
    "OH", "OK", "OR", "PA", "PR", "RI", "SC", "SD",
    "TN", "TX", "TT", "UT", "VT", "VI", "VA", "WA", "WV", "WI", "WY"
  )
)

# Print the data frame
print(state_data)
```

```{r}
# Calculate the count of victims for each state
state_counts <- table(dataset$state)

# Convert the counts to a data frame
state_data <- data.frame(state = names(state_counts), count = as.numeric(state_counts))

# Sort the data frame by count in descending order
state_data <- state_data[order(state_data$count, decreasing = TRUE), ]

# Create a horizontal bar chart
ggplot(state_data, aes(x = count, y = reorder(state, -count))) +
  geom_bar(stat = "identity", fill = "#CC0000") +
  geom_text(aes(label = count), hjust = -0.25) +
  labs(
    title = "States with Victims Count",
    x = "Number of Victims",
    y = "State"
  )

```

```{r}

library(stringr)
usa <- map_data("state")

usa_states <- usa %>%
  mutate(state = state.abb[match(str_to_title(usa$region), state.name)])

shoot_states <- dataset %>%
  group_by(state) %>%
  summarise(count = n())

usa_states_shoot_map <- merge(usa_states, shoot_states, by.x = "state", by.y = "state", sort = FALSE)
usa_states_shoot_map <- usa_states_shoot_map[order(usa_states_shoot_map$order),]

# Calculate the center of each state
state_centers <- usa_states_shoot_map %>%
  group_by(state) %>%
  summarize(center_long = mean(long), center_lat = mean(lat))

# Create the map with border lines and color
ggplot() +
  geom_polygon(data = usa_states_shoot_map, aes(x = long, y = lat, group = group, fill = count), color = 'black') +
  geom_path(data = usa_states_shoot_map, aes(x = long, y = lat, group = group), color = 'black') +
  geom_text(data = state_centers, aes(x = center_long, y = center_lat, label = state), hjust = 0.5, vjust = 0.5) +
  coord_map("polyconic") +
  scale_fill_continuous(low = "lightblue", high = "red") +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle('United States Shootings Heat Map')

```

# Racial Bias in US Fatal Police Shootings

```{r}
dataset$year <- as.integer(format(as.Date(dataset$date), "%Y"))
dataset %>%
  filter(year < 2021) %>%
  group_by(race, year) %>%
  mutate(race_killed = n()) %>%
  ggplot(aes(x = factor(year), y = race_killed, fill = race)) +
  labs(
    title = "Number of different victims of each race and each year:",
    x = "Year",
    y = "Count of Victims",
    fill = "Race:"
  ) +
  geom_col(position = 'dodge') +
  scale_fill_manual(values = my_colors) +
  scale_x_discrete(labels = c("2015", "2016", "2017", "2018", "2019", "2020"))  # Set custom x-axis labels

```

```{r}
race_pop_data <- read.csv("Race_Pop.csv")
race_pop_data
```

```{r}

race_pop_data <- read.csv("Race_Pop.csv")
# Extract the year from the date column
dataset$year <- as.integer(format(as.Date(dataset$date), "%Y"))

# Group and count victims by year and race, and set `.groups` to 'drop'
victim_counts <- dataset %>%
  group_by(year, race) %>%
  summarize(count = n(), .groups = 'drop')

# Merge the victim counts with population data
merged_data <- inner_join(victim_counts, race_pop_data, by = c("year", "race"))

# Calculate the proportion of victims relative to the population
merged_data <- merged_data %>%
  mutate(proportion = count / population) 

# Create the bar chart with explicit x-axis breaks
ggplot(merged_data, aes(x = year, y = proportion, fill = race)) +
  geom_col(position = 'dodge') +
  labs(
    title = "Proportion of Victims by Race and Year (Relative to Population)",
    x = "Year",
    y = "Proportion of Victims",
    fill = "Race"
  ) +
  scale_fill_manual(values = my_colors,) +
  scale_x_continuous(breaks = unique(merged_data$year))

```

# Classification by Age:-

## Box Plot of Victims by Age

```{r}
dataset %>%
  ggplot(aes(race, age)) +
  geom_boxplot(aes(fill = race)) +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Box Plot of Victims of Different Races by Age",
    x = "Race of Victim",
    y = "Age of Victim",
    fill = "Race:"
  )

# Calculate summary statistics by race
summary_stats <- dataset %>%
  group_by(race) %>%
  summarize(
    Q1 = quantile(age, 0.25),
    Q2 = median(age),
    Q3 = quantile(age, 0.75),
    IQR = IQR(age),
    Max = max(age),
    Min = min(age)
  )

# Print the summary statistics
print(summary_stats)

ggplot(dataset, aes(y = age, fill = "Ages")) +
  geom_boxplot() +
  labs(
    title = "Box Plot of Victim Age",
    y = "Age of Victim"
  ) +
  theme_minimal() +
  scale_fill_manual(values = "orange")

# Calculate summary statistics
summary_stats <- dataset %>%
  summarize(
    Q1 = quantile(age, 0.25),
    Q2 = median(age),
    Q3 = quantile(age, 0.75),
    IQR = IQR(age),
    Max = max(age),
    Min = min(age)
  )
print(summary_stats)

```

## Box Plot of Age of Victims Carrying Top 5 Weapons

```{r}
# Identify the top 5 weapons in the dataset
top_weapons <- dataset %>%
  group_by(armed) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  top_n(5) %>%
  pull(armed)

# Filter the dataset to include only the top weapons
filtered_dataset <- dataset %>%
  filter(armed %in% top_weapons)

# Create the box plot
boxplot_plot <- filtered_dataset %>%
  ggplot(aes(x = armed, y = age, fill = armed)) +
  geom_boxplot() +
  labs(
    title = "Box Plot of Age of Victims Carrying Top 4 Weapons and Unarmed",
    x = "Type of Armed Weapon",
    y = "Age of Victim",
    fill = "Type of Weapon"
  ) +
  scale_fill_manual(values = my_colors)

# Calculate and display summary statistics
summary_stats <- filtered_dataset %>%
  group_by(armed) %>%
  summarize(
    Q1 = quantile(age, 0.25),
    Q2 = median(age),
    Q3 = quantile(age, 0.75),
    IQR = IQR(age),
    max_outlier = Q3 + 1.5 * IQR,
    min_outlier = Q1 - 1.5 * IQR
  )

# Display summary statistics

boxplot_plot
summary_stats
```

# Consistencies in Police Shootings

### Cumulative Count of Victims Over Months (by Year)

```{r}
# Convert the 'date' column to a date format if it's not already
dataset$date <- as.Date(dataset$date)

# Extract year and month from the 'date' column
dataset$year <- format(dataset$date, "%Y")
dataset$month <- format(dataset$date, "%m")

# Filter the dataset for years 2015 to 2020
filtered_data <- dataset %>%
  filter(year >= "2015" & year <= "2020")

# Calculate cumulative victims
cumulative_data <- filtered_data %>%
  group_by(year, month) %>%
  summarize(cumulative_victims = n()) %>%
  arrange(year, month) %>%
  group_by(year) %>%
  mutate(cumulative_victims = cumsum(cumulative_victims))

# Create the cumulative line plot
ggplot(cumulative_data, aes(x = month, y = cumulative_victims, color = year, group = year)) +
  geom_line() +
  labs(
    title = "Cumulative Victims Over Months (by Year)",
    x = "Month",
    y = "Number of Victims"
  ) +
  scale_x_discrete(breaks = unique(cumulative_data$month), labels = month.name) +
  scale_color_discrete() +
  theme_minimal()

```

```{r}
# Filter the dataset for years 2015 to 2020
filtered_data <- dataset %>%
  filter(year >= 2015, year <= 2020)

# Calculate the number of victims for each year
victim_counts <- filtered_data %>%
  group_by(year) %>%
  summarize(total_victims = n())

victim_counts

```

### Total Victims by Year vs Signs of Mental Illness Status

```{r}
# Filter the dataset to include records up to 2020
filtered_dataset <- subset(dataset, year <= 2020)

# Extract the year from the date column
filtered_dataset$year <- as.integer(format(as.Date(filtered_dataset$date), "%Y"))

# Create a summary table of total victims by year and signs_of_mental_illness status
summary_data <- filtered_dataset %>%
  group_by(year, signs_of_mental_illness) %>%
  summarize(total_victims = n(), .groups = "drop")

# Create a bar graph
ggplot(summary_data, aes(x = factor(year), y = total_victims, fill = factor(signs_of_mental_illness))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Total Victims by Year vs Signs of Mental Illness Status",
    x = "Year",
    y = "Number of Victims",
    fill = "Signs of Mental Illness"
  )
summary_data

```

### Total Victims by Year vs Police Body Camera Status

```{r}

# Filter the dataset to include records up to 2020
filtered_dataset <- subset(dataset, year <= 2020)

# Extract the year from the date column
filtered_dataset$year <- as.integer(format(as.Date(filtered_dataset$date), "%Y"))

# Create a summary table of total victims by year and body_camera status
summary_data <- filtered_dataset %>%
  group_by(year, body_camera) %>%
  summarize(total_victims = n(), .groups = "drop")

# Create a bar graph
ggplot(summary_data, aes(x = factor(year), y = total_victims, fill = factor(body_camera))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Total Victims by Year vs Police Body Camera Status",
    x = "Year",
    y = "Number of Victims",
    fill = "Body Camera Status"
  )
summary_data
```

# Bibliography

US Police Shooting Data-set :-

<https://www.washingtonpost.com/graphics/investigations/police-shootings-database/>

Two-Letter State and Territory Abbreviations:-

<https://www.faa.gov/air_traffic/publications/atpubs/cnt_html/appendix_a.html>

US Population By Race:-

<https://www.census.gov/quickfacts/fact/table/US/PST045222>
